A forest is more than its trees: haplotypes and inferred ARGs

### 2024-09-03 paper/writeup.tex
new algorithm description

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

@@

\subsection{Extend Edges}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
    If neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
    suppose there exists a path which contains a node $3$,
    with parent and child nodes $4$ and $0$, respectively.
    If there is an edge between the nodes $4$ and $0$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
    This action reduces the length of $4 \to 0$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.
    With algorithm \ref{alg:edge} we only extend existing edges
    \comment{Note possible change pending algorithm change}
    over a larger interval on the genome,
    and remove unnecessary edges in the process.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{A visualization of the \textit{extend edges} method. Given two neighboring trees $T_1$ and $T_2$, the tree $T_1$ has a short path of edges $4 \to 3\to0$. The next tree contains a single edge $4 \to 0$. If node $3$ is not in $T_2$, then we will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$. 
    \label{fig:extending_diagram}
}
\end{figure}

@@

\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
{\BlankLine
	$\T = \left(T_{|\T|},..., T_1\right)$\;}
}
\Indm \For{$T_k\in \T$}{
	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
	\For{edges $f$ in $T_{k+1}$}{
		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
		Remove $f$ from $T_{k+1}$.\;
		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
		Update $\T$.\;
	}
	}
}\;						
\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
		}
}
\KwRet{$\T'$}\;
\end{algorithm2e}

@@

 \Fn{Shared Spans($T_1,T_2$)}{
 \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
 		Define $a_{ij}=|L_i\cap L_j|$.
 	}
 	\Else{Define $a_{ij}=0$.}
 Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
 }
 \KwRet{$A$}
 }
\BlankLine
\Fn{Discrepancy ($T_1,T_2$)}{
\For{each node $i\in N_{T_1}$}{
	Find nodes in $N_{T_2}$ which have the highest matching span.\;
	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
	Define $\alpha_i = a_{ij}$.\;
	\BlankLine
}
\begin{left}
$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
\BlankLine\BlankLine
\begin{left}
$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
\BlankLine\BlankLine
\KwRet{tree discrepancy, rmse}
}
\end{algorithm2e}

@@

@@

\begin{figure}
	\includegraphics[width=0.9\textwidth]{Edge_reduc_comp_speed.pdf}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}

@@

\begin{figure}
    \caption{
        Maybe: summaries of lengths and/or numbers of correctly and incorrectly added edges?
        (See notebook.)
        \label{fig:results_edges}
    }
\end{figure}

>>>>

% NOTE: these are in alphabetical order for now; we can revisit that ordering
%\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

@@

\subsection{Extend Haplotypes}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merge them, given the correct condition.
    For example, if neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree
     \ref{fig:extending_diagram}.
     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
     and denoted $EO$ and $EI$. \comment{Need better notation. Currently cant think of anything better.}
     For each edge $e \in EO$ with child $c$,
     we construct a path $p_c^{EO}$ comprised of edges in the $EO$ subforest.
     If there exists an edge $e'\in EI$ also with child $c$,
     we construct path $p_c^{EI}$ from edges in the $EI$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{EO}\cap p_c^{EI})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EO}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $EI$, see Figure \ref{complicated extending diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
%    We do so by identifying short paths (2 edges) in a tree.
%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
%    If neighboring trees contain edges with 
%    identical ancestor (parent) and descendant (child) nodes, 
%    we assume that there is implied coalescence of haplotypes between nodes, 
%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
%    suppose there exists a path which contains a node $3$,
%    with parent and child nodes $4$ and $0$, respectively.
%    If there is an edge between the nodes $4$ and $0$ in trees 
%    $T_{k+1}$ (or $T_{k-1}$),
%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
%    This action reduces the length of $4 \to 0$ on the genome,
%    and in some cases, completely removes the edge. 
%    We now perform this action on all such ancestral paths 
%    over the entire tree sequence.
%    With algorithm \ref{alg:edge} we only extend existing edges
%    \comment{Note possible change pending algorithm change}
%    over a larger interval on the genome,
%    and remove unnecessary edges in the process.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $EO$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $EI$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
	\end{center}
	\caption{More complicated extension}
	\label{fig:complicated_extending_diagram}
\end{figure}

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%

	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
	\SetKwFunction{ex}{Extend Paths}
	\SetKwFunction{fp}{Find Paths}
	\DontPrintSemicolon
	\LinesNotNumbered
	%\SetAlgoNoEnd
	\input{A tree sequence $\T$}
	\output{New tree sequence,$\T'$}
	\BlankLine
	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
	{
		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
		{\BlankLine
			$\T = \left(T_{|\T|},..., T_1\right)$\;
		}
	}
		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
		Compute subforests $EO\subseteq T_k$ and $EI\subseteq\in T_{k+1}$.\;
		\For{edges $e$ in $EO$}{
			Let $c$ be child node.\;
			$p_c^{EO} = [c]$\;
			$p_c^{EI} = [c]$\;
			$\bar{p}_c = []$ Path to be inserted\;
			\While{$\exists f\in EO$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
				$p = \tn{parent}(f)$\;
				Append $p$ to $p_c^{EO}$\;
				$c = p$\;
			}
			\While{$\exists f\in EI$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
				$p =\tn{parent}(f)$\;
				Append $p$ to $p_c^{EI}$\;
				$c = p$\;
			}
			\If{($p_c^{EO}\cap p_c^{EI})\backslash\{c\}\neq\emptyset$}{
				Merge paths up to intersection point\;
				$\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EI}_c$\;
				and sort by node time.\;
			}
			Insert $\bar{p}_c$ into $T_{k+1}$\;		
		}
	}\;
	Update $\T$.\;
	\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fp(fowards=\tn{True})\;
		$\T' \gets \T$.\fp(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
	}
	\KwRet{$\T'$}\;
}
\end{algorithm2e}

@@

\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
{	
\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
{\BlankLine
	$\T = \left(T_{|\T|},..., T_1\right)$\;}
}
	\For{$T_k\in \T$}{
	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
	\For{edges $f$ in $T_{k+1}$}{
		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
		Remove $f$ from $T_{k+1}$.\;
		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
		Update $\T$.\;
	}
	}
}\;						
\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
		}
}
\KwRet{$\T'$}\;
\end{algorithm2e}

@@

 \Fn{Shared Spans($T_1,T_2$)}{
 \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
 		Define $a_{ij}=|L_i\cap L_j|$.
 	}
 	\Else{Define $a_{ij}=0$.}
 Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
 }
 \KwRet{$A$}
 }
\BlankLine
\Fn{Discrepancy ($T_1,T_2$)}{
\For{each node $i\in N_{T_1}$}{
	Find nodes in $N_{T_2}$ which have the highest matching span.\;
	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
	Define $\alpha_i = a_{ij}$.\;
	\BlankLine
}
$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
\BlankLine\BlankLine
$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
\BlankLine
$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
\BlankLine\BlankLine
\KwRet{tree discrepancy, rmse, true proportion}
}
\end{algorithm2e}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion }
	\end{center}
	\label{fig:discrepancy-over-sample}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion}
	\end{center}
\end{figure}

@@

Notes on trapped unary spans:
Simulate a TS with the following parameters:
$msprime.sim_ancestry(1000, sequence_length=5e7, recombination_rate=1e-8, population_size=1e4, coalescing_segments_only=False, random_seed=1)$
From the above output we see that ratio of unary span bordering coalescing between 
- ExtendPaths/Truth = 0.4917
- ExtendEdges/Truth = 0.27828

The trapped unary span of our ground truth is 49\% of the TS.

Total unary span bordering coalescing: 255151996720.0 (51.0\%)
Total trapped unary span: 245038231157.0 (49.0\%)
Total coalescing span: 99950000000.0
Total unary span bordering coalescing: 125466159245.0 (100.0\%)
Total trapped unary span: 0.0 (0.0\%)
Total coalescing span: 99950000000.0

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
	\caption{Correct Unary Span for Extend Haplotypes (log)}
	\label{fig:correct-unary-span-log}
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
	\caption{Correct Unary Span for Extend Haplotypes}
	\label{fig:correct-unary-span}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/SL_vs_Num_edges_subplot.pdf}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL.pdf}
    \caption{
        Maybe: summaries of lengths and/or numbers of correctly and incorrectly added edges?
        (See notebook.)
        \label{fig:results_edges}
    }
\end{figure}
